Efficiently stable presentations from error-correcting codes

### 2023-10-03 efficient_stability.tex
cleared a bunch of comments

\begin{abstract}
We introduce a notion of \emph{efficient stability} for finite presentations of groups. Informally, a finite presentation using generators $S$ and relations $R$ is \emph{stable} if any map from $S$ to unitaries (say, in some finite von Neumann algebra) that approximately satisfies the relations (in the tracial norm) is close to the restriction of a representation of $G$ to the subset $S$. This notion and variants thereof have been extensively studied in recent years, in part motivated by connections to property testing in computer science. The novelty in our work is the focus on \emph{efficiency} which, informally, places an onus on small presentations (of size, say, logarithmic in the group size) and aims to achieve non-trivial tradeoffs between it and the stability---as, intuitively, more generators and relations place more constraints on the approximate homomorphism, making stability easier to achieve---whereas we wish to minimize the number of variables and constraints. \hnote{This is pretty wordy for an abstract... sometime later this week I can try to shorten it if someone hasn't already done it.}\mnote{I like the abstract as is...}

@@

A finer property which concerns us here is the \emph{soundness} of the code, a parameter that is connected to the notion of \emph{local testability}. A code can be (non-uniquely) specified through a \emph{parity-check matrix} $h\in\F^{m\times n}$ as $\code = \ker h$. The rows of $h$ are thought of as constraints (``parity checks'') that specify $\code$ as a subspace of $\F^n$. A code is called locally testable \emph{with soundness $\rho$} if for every $x\in \F^n$, $\frac{1}{m}|hx|\geq \rho \, \frac{1}{n}\, d(x,\code)$ \hnote{added the divide by $n$}, where $d(x,\code)$ denotes the minimum of $|x-c|$ over $c\in \code$.  Ideally one would like to design families of good codes such that in addition $\rho$ is bounded below by a constant independent of $n$.\footnote{Note that in principle a code can have a small minimal distance $d$, and still be locally testable with soundness $\rho>0$. So a family of codes can be ``locally testable'' without being ``good.'' \hnote{Should this be promoted to the main text? Feels a bit odd that it's a footnote.}\tnote{Yes, I moved some of it to the text in two different places. It was doing a couple things at once. The remaining part is less important.} } This is a challenging task, and the construction of families of good locally testable codes was only achieved very recently~\cite{LTC_DELLM}.

@@

\begin{definition}[Stability]\label{def:eff-stab-intro}
Let $G = \langle S:R\rangle $ be a finitely presented group, $\mu_S$ a distribution on $S$ and $\mu_R$ a distribution on $R$. For $\delta:[0,1]\to[0,1]$ such that $\lim_{t\to 0}\delta(t)=0$ and an integer $d\geq 1$ we say that the presentation $G=\langle S:R\rangle$ is $(\delta,\mu_S,\mu_R,d)$-stable if for every $(\eps,\mu_R)$-almost homomorphism $\phi: \mF(S) \to \mU(\C^d)$ there is a unitary representation $\psi: G \to \mU_d(\C)$ such that
\[ \Es{s\sim \mu_S} \big\|  \phi(s) - \psi(s) \big \|_f^2 \,\leq\, \delta(\eps)\;.\]
We refer to any function $\delta$ satisfying the above as a ``modulus of stability'' of the presentation.
\end{definition}

@@

We can now ask the question: is $G(h)$, the group presentation defined in~\eqref{eq:gh-intro}, stable according to Definition~\ref{def:eff-stab-intro}? One can verify that with $\mu_R$ and $\mu_S$ the uniform distribution on $R$ and $S$ respectively, $G(h)$ is $(\delta,\mu_S,\mu_R,1)$-stable for $\delta=\eps/\rho$ if and only if $\code = \ker h$ is locally testable with soundness $\rho>0$ \hnote{spelled out relationship between $\delta,\rho$}. But what about higher-dimensional approximate representations? Are these also stable, or does one need to make further requirements on $\code$ beyond local testability? We do not yet have a comprehensive answer to these questions. However, the answer is not trivial. In particular, and perhaps surprisingly, the group $G(h)$ is not necessarily Abelian, nor finite: see Remark~\ref{rk:non-abelian}. \hnote{this seems like an indirect way to motivate why it's nontrivial... why does non-Abelian-ness mean it's difficult? is there a better way of justifying the non-triviality of this question?}

\paragraph{Efficient stability.}
Faced with the apparent difficulty of studying the general question \hnote{I'm not so convinced we've properly justified its difficulty... }\tnote{The point is that given a presentation, identifying the group is not easy; e.g. the word problem is undecidable and so we can't straightforwardly simplify the relations in a canonical form or whatnot. We could be lucky and be in a very special situation. The non-Abelian example justifies that at least some mildly non-trivial groups show up. It's true that we didn't give that much thought to the problem, so feel free to tone down the language if you prefer?}, it is time to refine our focus and formulate the question which we \emph{do} address. To start, let us explicitly note that 
while stability has previously (for the most part) been studied as a question about a group, our formulation makes it a question about a \emph{presentation} of the group. In particular it is known~\cite{gowers2017inverse,de2019operator} that for finite groups $G$ (which are the only groups we consider in this paper) the multiplication table presentation, which has $|G|$ generators for every group element and $|G|^2$ relations for every pairwise product, is $\delta(\eps)=C\eps$-stable \hnote{hmm... technically the notion of stability in GH is not the same as the one defined above, right?}\tnote{Isn't it? The only difference I see is their focus on keeping control of the dimension $d'$, which Michael calls ``strongly flexibly stable'' or something like this. We could add a comment about this later, but here we're pretty high level.} \hnote{There are also some syntactic differences: I think GH define approx rep'n to be for every relation, not just on average, so there's no $\mu_R,\mu_S$, etc.}\tnote{No, of course it's on average over all relations (of the multiplication table representation. I don't see what the other differences are?} for some constant $C$ that is independent of the group. This holds even for approximate representations in arbitrary tracial von Neumann algebras, as long as one allows the notion of flexible stability alluded to above.

@@

\begin{theorem*}[Main, informal]
For every integer $k\geq 1$ there is a presentation $\Z_2^k = \langle S_k:R_k\rangle$ such that $|S_k|,|R_k| = 2^{\poly\log(k)}$ and furthermore this presentation is $\delta(\eps)=\poly(\log k,\eps)$-stable. \hnote{are we assuming some default setting for $\mu_S,\mu_R,d$, etc?}\tnote{The theorem statement is existential. So, implicitly, ``there exists $\mu_S,\mu_R$''. It's not clear enough?} \hnote{It's more that we haven't defined $\delta$-stable... we've defined $(\delta,\mu_R,\mu_S,d)$ stable only.}
\end{theorem*}

>>>>

\begin{abstract}
We introduce a notion of \emph{efficient stability} for finite presentations of groups. Informally, a finite presentation using generators $S$ and relations $R$ is \emph{stable} if any map from $S$ to unitaries (say, in some finite von Neumann algebra) that approximately satisfies the relations (in the tracial norm) is close to the restriction of a representation of $G$ to the subset $S$. This notion and variants thereof have been extensively studied in recent years, in part motivated by connections to property testing in computer science. The novelty in our work is the focus on \emph{efficiency} which, informally, places an onus on small presentations (of size, say, logarithmic in the group size) and aims to achieve non-trivial tradeoffs between it and the stability---as, intuitively, more generators and relations place more constraints on the approximate homomorphism, making stability easier to achieve---whereas we wish to minimize the number of variables and constraints.

@@

A finer property which concerns us here is the \emph{soundness} of the code, a parameter that is connected to the notion of \emph{local testability}. A code can be (non-uniquely) specified through a \emph{parity-check matrix} $h\in\F^{m\times n}$ as $\code = \ker h$. The rows of $h$ are thought of as constraints (``parity checks'') that specify $\code$ as a subspace of $\F^n$. A code is called locally testable \emph{with soundness $\rho$} if for every $x\in \F^n$, $\frac{1}{m}|hx|\geq \rho \, \frac{1}{n}\, d(x,\code)$ \hnote{added the divide by $n$}, where $d(x,\code)$ denotes the minimum of $|x-c|$ over $c\in \code$.  Ideally one would like to design families of good codes such that in addition $\rho$ is bounded below by a constant independent of $n$.\footnote{Note that in principle a code can have a small minimal distance $d$, and still be locally testable with soundness $\rho>0$. So a family of codes can be ``locally testable'' without being ``good.''} This is a challenging task, and the construction of families of good locally testable codes was only achieved very recently~\cite{LTC_DELLM}.

@@

\begin{definition}[Stability]\label{def:eff-stab-intro}
Let $G = \langle S:R\rangle $ be a finitely presented group, $\mu_S$ a distribution on $S$ and $\mu_R$ a distribution on $R$. For $\delta:[0,1]\to[0,1]$ such that $\lim_{t\to 0}\delta(t)=0$ and an integer $d\geq 1$ we say that the presentation $G=\langle S:R\rangle$ is $(\delta,\mu_S,\mu_R,d)$-stable if for every $(\eps,\mu_R)$-almost homomorphism $\phi: \mF(S) \to \mU(\C^d)$ there is a unitary representation $\psi: G \to \mU_d(\C)$ \hnote{there's some mismatch, is it $\mU(\C^d)$ or $\mU_d(\C)$?} such that
\[ \Es{s\sim \mu_S} \big\|  \phi(s) - \psi(s) \big \|_f^2 \,\leq\, \delta(\eps)\;.\]
We refer to any function $\delta$ satisfying the above as a ``modulus of stability'' of the presentation.
\end{definition}

@@

We can now ask the question: is $G(h)$, the group presentation defined in~\eqref{eq:gh-intro}, stable according to Definition~\ref{def:eff-stab-intro}? One can verify that with $\mu_R$ and $\mu_S$ the uniform distribution on $R$ and $S$ respectively, $G(h)$ is $(\delta,\mu_S,\mu_R,1)$-stable for $\delta=\eps/\rho$ if and only if $\code = \ker h$ is locally testable with soundness $\rho>0$ \hnote{spelled out relationship between $\delta,\rho$}. But what about higher-dimensional approximate representations? Are these also stable, or does one need to make further requirements on $\code$ beyond local testability? We do not yet have a comprehensive answer to these questions. \hnote{elaborated a little on the challenges} However, the answer cannot be straightforward: even deducing basic properties about the group $G(h)$ given its presentation -- let alone its stability properties -- appears to be a challenging task. For example, there are examples of parity check matrices $h$ for which the group $G(h)$ is non-Abelian or even infinite; see Remark~\ref{rk:non-abelian}.

\paragraph{Efficient stability.}
Faced with the apparent difficulty of studying the general question, it is time to refine our focus and formulate the question which we \emph{do} address. To start, let us explicitly note that 
while stability has previously (for the most part) been studied as a question about a group, our formulation makes it a question about a \emph{presentation} of the group. In particular it is known~\cite{gowers2017inverse,de2019operator} that for finite groups $G$ (which are the only groups we consider in this paper) the multiplication table presentation, which has $|G|$ generators for every group element and $|G|^2$ relations for every pairwise product, is $\delta(\eps)=C\eps$-stable\footnote{\hnote{added this:}Here we assume that the $\mu_R,\mu_S$ are uniform over the relations and generators of the multiplication table presentation, respectively. Furthermore, \cite{gowers2017inverse,de2019operator} use a notion of flexible stability, where the nearby exact representation may act on a larger dimension.} for some constant $C$ that is independent of the group. This holds even for approximate representations in arbitrary tracial von Neumann algebras, as long as one allows the notion of flexible stability alluded to above.

@@

\begin{theorem*}[Main, informal]
For every integer $k\geq 1$ there is a presentation $\Z_2^k = \langle S_k:R_k\rangle$ such that $|S_k|,|R_k| = 2^{\poly\log(k)}$ and furthermore this presentation is $(\delta,\mu_{S_k},\mu_{R_k},d)$-stable for all $d \in \N$, where $\delta(\eps)=\poly(\log k,\eps)$, $\mu_{S_k}$ is uniform over the generators and $\mu_{R_k}$ is some distribution over the relations. \hnote{Spelled the conditions out because if we just let $\mu_S,\mu_R$ be anything, then we get a trivial statement if $\mu_S$, say, is just supported on one generator and $\mu_R$ is just on one relation.}
\end{theorem*}
